Typed Interactions for NLU in Opal

### 2017-12-13 paper.tex
Edits/revisions

Today, computers are our personal assistants. They are integrated with our lives
and they help everyday people do everyday tasks. Most people who use computers
today never touch a command shell, and even those who do would often rather not.
Graphical interfaces are amazingly powerful, and are often good enough to allow
users to easily interact with the computer, but even those have their drawbacks.
We would really like our personal assistant to respond to commands in human
language:
\begin{itemize}
\item ``Hey Siri! Play some music.''
\item ``Ok Google. Navigate to my hotel.''
\item ``Slackbot, set my status to active.''
\end{itemize}
This is the realm of Natural Language Understanding (NLU).\footnote{We do not
  distinguish between spoken or written language when talking about NLU in
  general.}

@@

\subsection{An Example}
Consider a simple reminder interface to a calendar or other scheduling
assistant. A user might ask the application to set a reminder containing a
message and a time (for simplicity, we will just allow the user to specify
\ts{later} or \ts{tomorrow}) or she might ask the application to get any active
reminders. Since Opal is embedded in TypeScript,\cite{typescript} a typical Opal
application might express this interface with the following declarations:
\begin{minted}{typescript}
  type Msg = string;
  type Time = "later" | "tomorrow";
  type Set = {
    msg: Msg,
    time: Time
  };
  type Intent =
    | { tag: "Set", data: Set }
    | { tag: "Get", data: {} };
\end{minted}

Since we would like to interact with this application via human language, we
need some way of converting a user \emph{utterance} to something of type
\ts{Intent}. Figure \ref{fig:drycleaning} shows the desired
data representation of a user utterance.

@@

\subsection{Search Strategies}
When entities are declared, the user must specify a ``search strategy.'' In the
case of Wit, there are three main strategies:
\begin{description}
\item[Free Text] Entities that are found via free text matching are just strings
of text from the utterance. In our example, the message is a free text entity.
\item[Keywords] Keywords entities are just words from some pre-defined set.
  Things like \ts{later} and
  \ts{tomorrow} would be captured via keyword searching.
\item[Trait] These are slightly more abstract. Traits capture something that is
  a quality of the utterance as a whole. Something like a user's intent or even
  mood could be a trait.
\end{description}

@@

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
declarations from Section \ref{introduction} written in our DSL.

@@

The main thing to notice is that the keyword \ts{type} has been replaced by
entity declarations \ts{free-text}, \ts{keywords}, and \ts{trait}, as well as
\ts{alias}. There are a number of benefits of entity declarations. First and
foremost, it allows the user to explicitly define the NLU engine configuration
alongside the program types. With types defined in our DSL, we can automatically
generate configuration for entities that is guaranteed to be consistent with the
types in the program. When a \ts{Time} is returned to the user, it will be
guaranteed to be either \ts{"later"} or \ts{"tomorrow"}. This alone solves a lot
of the problems that tend to come up when writing these types of applications.

Another benefit of these bespoke declaration forms is simplified syntax. When a
\ts{free-text} entity is declared, there is no need to write ``\ts{= string}'',
because that is implied by the search strategy. Similarly, since traits are
always tagged unions, we can write them in (almost) ML syntax, rather than
writing out the idiomatic TypeScript workaround.

@@

Figure \ref{fig:ast} shows the AST that is produced when we parse the DSL.
Looking at this, it also becomes clear that the limitations on the types allowed
in a trait declaration are fairly strict. As of writing, there are only three
types allowed within a trait declaration: previously defined entities, literals,
and records (where each element is subject to the same restrictions). These
limitations match up with the capabilities of NLU engines in general. While many
backends provide the ability to parse numbers, dates, and other data, these
types of entities are implementation specific. In Section \ref{future}, we
discuss plans to expand the allowed types in a modular way.

@@

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from Section \ref{description}. We begin by formalizing the type DSL
as shown in Figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{Type DSL}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \tau
    \rangle $$
    \caption{Relevant Opal Types}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{The core language.}
\end{figure}
There are similarities between these definitions and the ones in Figure
\ref{fig:ast}, but the core language has some slightly different goals. First,
we remove aliases and require that all type definitions are written ``in-line.''
Next, rather than split the definition by allowing tags within $\fcy{T}$, we
also require that entity definitions are inlined. This means that the data a
user expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---this
formulation can represent the same types as the actual system. This approach
simply gives up some usability to simplify reasoning.

In addition to a grammar for the core language itself, in Figure
\ref{fig:grammar_b} we define a grammar of types, $\tau$. We can think of $\tau$
as representing the final form of the data that a user hopes to obtain, with no
information about the entities that generated the data. They map almost directly
into types that are used by the underlying Opal system. We add a few types that
are noteworthy: the type {\bf string} is added to account for FreeText entities;
and $\langle {\tau} \rangle$ represents an arbitrary union, and covers the cases
of Keywords and Trait entities.

@@

One interesting thing to point out is that the record in the Trait declarations
becomes a union type. This makes sense, since a trait represents a property of
the utterance as a whole, and thus can decide the structure of the response.

@@

The \ff{lookup} function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free-text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities. The function \ff{cfg} takes a
declaration and produces a configuration for that declaration.
$$ \ff{cfg}\ :\ \fcy{D} \to \ff{Tag} \rightharpoonup \tau $$
We define \ff{cfg} in Figure \ref{fig:cfg}. The exact details of the else branch
of the Trait case are implementation specific, but the basic idea to map the
\ff{cfg} function over all nested declarations. This achieves the goal of
capturing every tag in the declaration.

@@

Now, we call a response well-formed with respect to $d : \fcy{D}$ if and only if
it returns appropriate data on \ff{lookup}. More formally, for some $r :
\ff{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{d} \iff \\
  &\ \ \forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(d, \etag{t})\ \text{defined} \Rightarrow  \\
  &\qquad \ff{lookup}(\etag{t}, r) : \ff{cfg}(d, \etag{t})
\end{align*}
We will use well-formedness as our basic assumption that the application has
been configured correctly relative to the response.

@@

In the event that $r$ is not well-formed with respect to the configuration
generated by $d$, $\ff{parse}(d, r)$ is undefined. Otherwise, we define
\ff{parse} by case analysis on the declaration.
\begin{align*}
  \ff{parse}(\text{FreeText}(\etag{t}), r) = \ff{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a {\bf string}, we can simply look up the value of the
appropriate entity from the response.
\begin{align*}
  \ff{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \ff{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from \ff{lookup}.
\begin{align*}
  \ff{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \ff{parseTy}(rec[\ff{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for \ff{parse} simply matches the value in the response to an
index into the record, and interprets that record entry using \ff{parseTy}. It
is important to note that since \ff{interp} returns a union here, we only need
to return a value that is in some branch of that union. In this case, we use the
\ff{lookup} value to decide which branch.

@@

\section{Implementation} \label{implementation}
\begin{figure*}
  \centering
  \begin{tikzpicture}[node distance = 3cm, auto]
    \node [block] (types) {Type DSL};
    \node [block, right of=types] (decls_hs) {Decl (hs)};
    \node [block, right of=decls_hs] (decls_ts) {Runtypes (ts)};
    \node [block, below of=decls_ts] (cfg) {NLU Config};
    \node [block, above of=decls_ts] (ts_types) {Types (ts)};
    \node [cloud, right of=decls_ts] (ts) {Opal};
    \node [cloud, right of=cfg] (backend) {Wit};

    \path [line] (types) -- (decls_hs);
    \path [line] (decls_hs) -- (decls_ts);
    \path [line] (decls_hs) -- (cfg);
    \path [line] (decls_hs) -- (ts_types);
    \path [line] (decls_ts) -- (ts);
    \path [line] (ts_types) -- (ts);
    \path [line] (cfg) -- (backend);
  \end{tikzpicture}
  \caption{The configuration process.}
  \label{fig:process}
\end{figure*}
A high-level workflow for our implementation is shown in Figure
\ref{fig:process}. The programmer writes types in our DSL, and uses our Haskell
tool to parse the declarations. The tool serializes three artifacts: a set of
TypeScript type declarations, a runtypes datastructure, and a set of JSON
configuration files for Wit. The programmer simply imports the types and
runtypes into their Opal project, uploads the JSON files to Wit, and the system
is ready to go.

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool turns
code in the type DSL into TypeScript code and JSON configuration, and ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

@@

\subsection{Runtime Library}
At run-time typical application using NLU will go through a fairly consistent
cycle.
\begin{enumerate}
\item Ask user for input, converting voice to text if necessary.
\item Query the NLU engine and obtain a response for the given input.
\item Parse the response into application data.
\item Process parsed data, and repeat.
\end{enumerate}
Steps 1 and 4 are generally application-specific, so we have elected to provide
a library which handles steps 2 and 4. For each backend that Opal
supports,\footnote{At the time of writing, Wit is the only supported backend.}
we provide utilities for querying the API and parsing the response data. The
network code involved in obtaining the response is fairly uninteresting, so we
will focus on the parsing,

@@

The other difference arises because Opal does not have dependent types. We
cannot completely faithfully transcribe the parse function from above, because
it is not obvious that it can be typed at all. The solution here is to combine
Opal's dynamic typing with runtypes' dynamic checking feature to create a
function that appears to be dependently typed. First, we use dynamic types
within the parse function. While building the response data, we assume, where
necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
runtypes object to check the parsed object, and we cast it to the desired type.
The definition is shown in Figure \ref{fig:parse}. The cast is safe, so long as
the user respects the invariant that parse is always called with compatible type
and runtype parameters.

@@

\subsection{Type Expressiveness}
The first important thing we found when starting to write types in our DSL is
that the standard three entity types are not quite expressive enough for real
applications. For example, when trying to implement a small chatbot for
scheduling calendar meetings, we wanted to write something like,
\begin{minted}{typescript}
  free-text Person
  ...
  trait Intent = <Schedule> {
    person: Person,
    date: Date,
    time: Time
  }
\end{minted}
but we found that we had no way to express a date or a time. Technically
speaking, it might have been possible to encode dates and times as free-text and
do a second parsing pass in the application code, but that solution is far from
elegant. We discuss one potential solution to this problem in Section
\ref{future>types}.

@@

It turns out that this might actually be a good thing, if viewed from the
perspective of the NLU engine itself. While \ts{Absolute} and \ts{Relative} are
the same intent from a philosophical perspective, they might be parsed very
differently. The utterance
\begin{center}
  \emph{Schedule my Wednesday meeting with Chris.}
\end{center}
is very different in structure from
\begin{center}
  \emph{Set a meeting with James tomorrow.}
\end{center}
Thus, it might be the case that forcing the programmer to write slightly
different types will make it much easier for the NLU backend to find the correct
information.

Overall, there is likely some work to be done in terms of making our types more
idiomatic, but for now we err on the side of making life easier for the NLU
engine.

@@

\subsubsection{More Types} \label{future>types}
In addition to the three ``universal'' kinds of entities (free-text, keywords,
and traits), the Wit NLU engine provides more specialized ways of looking up
entities. Wit provides support for parsing numbers, dates, times, and other
specific forms of text. We elected to keep things simple at first and exclude
these built-in entities, but adding support for them seems like a very
reasonable next-step.

>>>>

Today, computers are our personal assistants. They are integrated with our lives
and they help everyday people do everyday tasks. Most people who use computers
today never touch a command shell, and even those who do would often rather not.
Graphical interfaces are amazingly powerful, and are often sufficient to allow
users to easily interact with the computer, but even those have their drawbacks.
We would really like our personal assistant to respond to commands in human
language:
\begin{itemize}
\item ``Hey Siri! Play some music.''
\item ``Ok Google. Navigate to my hotel.''
\item ``Slackbot, set my status to active.''
\end{itemize}
This is the realm of Natural Language Understanding (NLU).\footnote{We do not
  distinguish between spoken or written language when talking about NLU in
  general.}

@@

\subsection{An Example}
Consider a simple \emph{reminders} interface to a calendar or other scheduling
assistant. A user might ask the application to set a reminder containing a
message and a time (for simplicity, we will just allow the user to specify
\ts{later} or \ts{tomorrow}) or she might ask the application to get any active
reminders. Since Opal is embedded in TypeScript,\cite{typescript} a typical Opal
application might express this interface with the following declarations:
\begin{minted}{typescript}
  type Msg = string;
  type Time = "later" | "tomorrow";
  type Set = {
    msg: Msg,
    time: Time
  };
  type Intent =
    | { tag: "Set", data: Set }
    | { tag: "Get", data: {} };
\end{minted}

We would like to interact with this application via human language, so we need
some way of converting a user's \emph{utterance} to something of type
\ts{Intent}. Figure \ref{fig:drycleaning} shows the desired data representation
of a user utterance.

@@

\subsection{Search Strategies}
When entities are declared, the user must specify a ``search strategy.'' In the
case of Wit, there are three main strategies:
\begin{description}
\item[Free Text] Entities that are found via free text matching are just strings
of text from the utterance. In our example, the message is a free text entity.
\item[Keywords] Keywords entities are just words from some pre-defined set.
  Things like \ts{later} and
  \ts{tomorrow} would be captured via keyword searching.
\item[Trait] These are slightly more abstract. Traits capture something that is
  a quality of the utterance as a whole. Something like a user's intent or even
  their mood could be a trait.
\end{description}

@@

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Figure \ref{fig:dsl_decls} shows the
same type declarations from Section \ref{introduction} written in our DSL.

@@

The main thing to notice is that the keyword \ts{type} has been replaced by
entity declarations \ts{free-text}, \ts{keywords}, and \ts{trait}, or by
\ts{alias}. There are a number of benefits of entity declarations. First and
foremost, it allows the user to explicitly define the NLU engine configuration
alongside the program types. With types defined in our DSL, we can automatically
generate configuration for entities that is guaranteed to be consistent with the
types in the program. When a \ts{Time} is returned to the user, it will be
guaranteed to be either \ts{"later"} or \ts{"tomorrow"}. This alone solves a lot
of the problems that tend to come up when writing these types of applications.

Another benefit of these bespoke declaration forms is simplified syntax. When a
\ts{free-text} entity is declared, there is no need to write ``\ts{= string},''
because that is implied by the search strategy. Similarly, since traits are
always tagged unions, we can write them in (almost) ML syntax, rather than
writing out the standard TypeScript workaround.

@@

Figure \ref{fig:ast} shows the AST that is produced when we parse the DSL.
Looking at this, it also becomes clear that the limitations on the types allowed
in a trait declaration are fairly strict. As of writing, there are only three
types allowed within a trait declaration: previously defined declarations,
literals, and records (where each element is subject to the same restrictions).
These limitations match up with the capabilities of NLU engines in general.
While many backends provide the ability to parse numbers, dates, and other data,
these types of entities are implementation specific. In Section \ref{future}, we
discuss plans to expand the allowed types in a modular way.

@@

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from Section \ref{description}. We begin by writing down a core
language for the type DSL as shown in Figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{The core language.}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \tau
    \rangle $$
    \caption{Opal types.}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{Formal structures.}
\end{figure}
There are similarities between these definitions and the ones in Figure
\ref{fig:ast}, but the core language has some slightly different goals. First,
we remove aliases and require that all type definitions are written ``in-line.''
Next, rather than split the definition by allowing tags within $\fcy{T}$, we
also require that entity definitions are inlined. This means that the data a
user expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---this
formulation can represent the same types as the actual system. This approach
simply gives up some usability to simplify reasoning.

In addition to a grammar for the core language itself, in Figure
\ref{fig:grammar_b} we define a grammar of types, $\tau$. We can think of $\tau$
as representing the final form of the data that a user hopes to obtain, with no
information about the entities that generated the data. They map almost directly
into types that are used by the underlying Opal system. We add a few types that
are noteworthy. The type {\bf string} is added to account for FreeText entities.
The type $\langle {\tau} \rangle$ represents an arbitrary union and covers the
cases of Keywords and Trait entities.\footnote{This notation is slightly
unconventional, but we prefer to write $\langle \tau \rangle$ or $\langle \tau_1
\mid \dots \mid \tau_n \rangle$ rather than simply $\tau_1 \mid \dots \mid
\tau_n$ to separate it from the surrounding syntax.}

@@

One interesting thing to point out is that the record in the Trait declarations
becomes a union type. This makes sense, since a trait represents a property of
the utterance as a whole, and thus can decide the structure of the rest of the
response.

@@

The \ff{lookup} function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities. The function \ff{cfg} takes a
declaration and produces a configuration for that declaration.
$$ \ff{cfg}\ :\ \fcy{D} \to \ff{Tag} \rightharpoonup \tau $$
We define \ff{cfg} in Figure \ref{fig:cfg}. The exact details of the else branch
of the Trait case are implementation specific, but the basic idea to map the
\ff{cfg} function over all nested declarations. This achieves the goal of
capturing every tag in the declaration.

@@

Now, we call a response well-formed with respect to $d : \fcy{D}$ if and only if
it returns appropriate data on \ff{lookup}. More formally, we say that for some
$r : \ff{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{d} \iff \\
  &\ \ \forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(d, \etag{t})\ \text{defined} \Rightarrow  \\
  &\qquad \ff{lookup}(\etag{t}, r) : \ff{cfg}(d, \etag{t})
\end{align*}
We will use well-formedness as our basic signal that a the application and the
response have been configured to be compatible.

@@

In the event that $r$ is not well-formed with respect to the configuration
generated by $d$, $\ff{parse}(d, r)$ is undefined. Otherwise, we define
\ff{parse} by case analysis on the declaration.
\begin{align*}
  \ff{parse}(\text{FreeText}(\etag{t}), r) = \ff{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free text
entity is just a {\bf string}, we can simply look up the value of the
appropriate entity from the response.
\begin{align*}
  \ff{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \ff{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from \ff{lookup}.
\begin{align*}
  \ff{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \ff{parseTy}(rec[\ff{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for \ff{parse} simply matches the value in the response to an
index into the record, and interprets that record entry using \ff{parseTy}. It
is important to note that since \ff{interp} returns a union here, we only need
to return a value that is in some branch of that union. In this case, we use the
\ff{lookup} value to decide which branch.

@@

\section{Implementation} \label{implementation}
\begin{figure*}
  \centering
  \begin{tikzpicture}[node distance = 3cm, auto]
    \node [block] (types) {Type DSL};
    \node [block, right of=types] (decls_hs) {Decl (hs)};
    \node [block, right of=decls_hs] (decls_ts) {AST (ts)};
    \node [block, below of=decls_ts] (cfg) {NLU Config};
    \node [block, above of=decls_ts] (ts_types) {Types (ts)};
    \node [cloud, right of=decls_ts] (ts) {Opal};
    \node [cloud, right of=cfg] (backend) {Wit};

    \path [line] (types) -- (decls_hs);
    \path [line] (decls_hs) -- (decls_ts);
    \path [line] (decls_hs) -- (cfg);
    \path [line] (decls_hs) -- (ts_types);
    \path [line] (decls_ts) -- (ts);
    \path [line] (ts_types) -- (ts);
    \path [line] (cfg) -- (backend);
  \end{tikzpicture}
  \caption{The configuration process.}
  \label{fig:process}
\end{figure*}
A high-level workflow for our implementation is shown in Figure
\ref{fig:process}. The programmer writes types in our DSL and uses our Haskell
tool to parse the declarations. The tool serializes three artifacts: a set of
TypeScript type declarations, an AST datastructure, and a set of JSON
configuration files for Wit. The programmer simply imports the types and AST
into their Opal project, uploads the JSON files to Wit, and the system is ready
to go.

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

@@

\subsection{Run-time Library}
At run-time, a typical application using NLU will go through a fairly consistent
cycle.
\begin{enumerate}
\item Ask user for input, converting voice to text if necessary.
\item Query the NLU engine and obtain a response for the given input.
\item Parse the response into application data.
\item Process parsed data, and repeat.
\end{enumerate}
Steps 1 and 4 are generally application-specific, so we have elected to provide
a library which handles steps 2 and 4. For each backend that Opal
supports,\footnote{At the time of writing, Wit is the only supported backend.}
we provide utilities for querying the API and parsing the response data. The
network code involved in obtaining the response is fairly uninteresting, so we
will focus on the parsing.

@@

The other difference arises because Opal does not have dependent types. We
cannot completely faithfully transcribe the \ff{parse} function from above,
because it is not obvious that it can be typed at all. The solution here is to
combine Opal's dynamic typing with Runtypes' dynamic checking feature to create
a function that appears to be dependently typed. First, we use dynamic types
within the \ff{parse}. While building a response object, we assume, where
necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
Runtypes object to check the parsed result, and we cast the result to the
desired type. The definition is shown in Figure \ref{fig:parse}. The cast is
safe, so long as the user respects the invariant that parse is always called
with compatible type and runtype parameters.

@@

\subsection{Type Expressiveness}
The first important thing we found when starting to write types in our DSL is
that the standard three entity types are not quite expressive enough for real
applications. For example, when trying to implement a small chatbot for
scheduling calendar meetings, we wanted to write something like,
\begin{minted}{typescript}
  free-text Person
  ...
  trait Intent = <Schedule> {
    person: Person,
    date: Date,
    time: Time
  }
\end{minted}
but we found that we had no way to express a date or a time. Technically
speaking, it might have been possible to encode dates and times as free text and
do a second parsing pass in the application code, but that solution is far from
elegant. We discuss one potential solution to this problem in Section
\ref{future>types}.

@@

It turns out that this might actually be a good thing, if viewed from the
perspective of the NLU engine itself. While \ts{Absolute} and \ts{Relative} are
the same intent from a philosophical perspective, they might be parsed very
differently. The utterance
\begin{center}
  \emph{Schedule my Wednesday meeting with Chris.}
\end{center}
is very different in structure from
\begin{center}
  \emph{Set a meeting with James tomorrow.}
\end{center}
Thus, it might be the case that forcing the programmer to write slightly
different types will make it much easier for the NLU backend to find the correct
information. There is likely some work to be done in terms of making our types
more idiomatic, but for now we err on the side of making life easier for the NLU
engine.

\subsection{Relative Success}
There is still a lot to be done on this work and in this space, but as of
writing, we feel that we have proven the concept fairly well. The system does
the job it intends to, and is usable for actual applications. In general, using
the tools feels fairly natural. The Haskell tool produces the output that a user
would expect, and the programming interface provides the functionality that a
user needs to actually write an application that uses NLU. Overall, we are happy
with our approach to the stated problem as well as with the results of the
implementation.

@@

\subsubsection{More Types} \label{future>types}
In addition to the three ``universal'' kinds of entities (free text, keywords,
and traits), the Wit NLU engine provides more specialized ways of looking up
entities. Wit provides support for parsing numbers, dates, times, and other
specific forms of text. We elected to keep things simple at first and exclude
these built-in entities, but adding support for them seems like a very
reasonable next-step.
